### 2025-04-08 tfp2025 3/review-response.md

> Thank you for your submission to the TFP post-proceedings. I am delighted to inform you that your paper has been accepted. I hope you find the reviews helpful in making final revisions. Please complete these by Weds 16th April. If you need an extra page or two, that's fine.
> 
> ----------------
>  REPORT
> ----------------
> 
> ## Synopsis
> 
> This paper describes parsers for a (subset of) context free languages,
> implemented in Agda, inspired by Elliott's recent work in this
> direction. The major hurdle to overcome is establishing such parsers are
> structurally recursive. The solution, as is used more widely in parsing
> of regular expressions, is to compute the derivative of a context free
> grammar -- and then perform induction on the input string.
> 
> ## Review
> 
> The paper is much improved from the submitted version. It now gives a
> fairly clean account of parsing using derivatives. Even though not all
> context free languages can be parsed in this fashion -- it is entirely
> plausible that this approach can be scaled to handle them. The main
> result, the correct parser in section 3.3 gives an extremely clean
> treatment of recursion.
> 
> I would like to offer a few suggestions on how the presentation and
> material may be improved.
> 
> ### Major suggestions
> 
> * I found the terminology used by the paper a bit confusing at
>  times. Languages are String -> Set; parsers decide languages – so much
>  is clear. But there is an (implicit) mix of (data) types and parsers -
>  why else would you want to have operations such as · operation in your
>  languages used? Similarly, the fragment of context free grammars
>  introduced in section 3 is called 'Desc' (suggesting it is a
>  description of a data type) – where I would expect this to be a (deep
>  embedding of a fragment of) context free grammars.

TODO: State somewhere that grammars and data types are tightly connected.

TODO: See if an example of naive ambiguous expressions could show the usefulness
of the scalar multiplication operation.

Sketch: data types are not that much different from grammars. Both are initial
algebras of endofunctors. The difference is the category on which they are
defined. I think it would be good to state this explicitly somewhere early on in
the paper.

Also, we should explain that we use fairly rigid terminology when we say
'languages', 'grammars', and 'parsers'. A *language* is an abstract concept that
can be described concretely in various ways, traditionally: by a set of strings,
by a grammar, by an automaton, etc. We stretch this definition somewhat with our
type theoretic approach. Instead of a set of strings (a function from strings to bool), we consider a function from strings to types to be a concrete representation of a language. The crucial difference is that a string can now be
"in" a language in multiple distinct ways, allowing us to model ambiguity.

Traditionally, a *grammar* is a list of nonterminals and collection of rewrite 
rules from the nonterminals to lists of nonterminals and terminals. A grammar
defines a language, namely the set of all strings (lists of nonterminals)
reachable by applying rewrite rules starting from a designated the starting
nonterminal. Note that grammars have more structure than languages. For example,
a string can be generated by a grammar in multiple different ways. Our type
theoretic definition of language correlates, in that respect, more directly to
traditional grammars. 
However, in this paper we define grammars in a slightly different way. We use an
algebraic/expression based definition ... TODO!

A *parser* is usually defined as a function from strings to booleans in theory
or a program which turns a string into a parse tree in practice.

> * One of the main technical contributions of the paper is the treatment
>  of the recursion and derivatives. To compute the derivative, there is
>  an additional mu constructor, introducing an indirection to
>  distinguish between the original language and the recursive calls to
>  the newly computed derivative. Given that the paper is mostly
>  interested in parsing, one might wonder if the indirection is strictly
>  necessary: couldn't the intended semantics (i.e. recursive call to the
>  derivative function of the _original_ grammar) not be passed as an
>  argument to δ₀? If this isn't the case, it might be worth
>  explaining/motivating why the direct solution does not work - I could
>  imagine it causes problems for the termination checker, for instance.

TODO: I think this indeed would not work out nicely, but we could explain why it
does not work in the paper. I don't think it is generally useful to explore
every possible alternative in the paper, but perhaps this seems obvious enough
to warrant some explanation.  Also, our data type is not closed under
derivatives

TODO: If you take a more big-step/recursive descent approach with memoization,
then it might be possible. We could explain this in the discussion of possible
future work.

TODO: At the end of Section 2 we should mention something about the switch from
big step/recursive descent to small step/derivatives.

> ### Minor suggestions and clarifications
> 
> * Remark 2 : (either here or in the future work) – what changes would be
>  necessary to enumerate all parse trees?

One of the main difficulties is that our nullability correctness theorem is
not a one-to-one equivalence. We intentionally exclude parses which take 
unproductive recursive iterations.

Furthermore, I don't think ambiguous grammars are useful in practice (in CS), so
our effort would be better spent developing better tools to check and prove
that our languages are not ambiguous, avoiding the whole issue of having
multiple parse trees.

> * One important reason for separating out nullability and derivatives is
>  that it becomes possible to define the parser by induction on the
>  input string. Otherwise, it is not so clear that parsing will
>  terminate at all – and often does not. I'd suggest clarifying this
>  point halfway through page 6.

Good point. Although, this is not obvious. I need some time to think about this.

> * Another reason to avoid · would be that it makes the Desc type
>  large... Removing it would allow Desc to live in Set - unless there
>  are particular applications, but these are not given in the paper.

The · is necessary for the derivative of sequencing of grammars. Well, you
could possibly replace is by a Sigma type, perhaps? Let me think about that.
We also use it in the derivative of characters, but that is easier to replace.

> * I found the ⋄ notation rather confusing... Is it worthwhile? The
>  definitions are not so long - why not inline them when introducing the
>  semantics of Desc?

It is clear that I at least need to spend some more time explaining ⋄.

> * The fixpoint operation on Desc to produce a Lang is not the standard
>  one - presumably because you want to define a predicate on strings
>  rather than a Set. Given that this is a key contribution - perhaps it
>  worth explaining a bit better.

The fixed point is really not that much different from the normal fixed point.
The only difference is that it passes along the input string `w`. I could try
to spend a paragraph explaining that.

I would like to say that this is the normal fixed point - just in a different
category - but I don't think that translates directly to this Agda definition.

> * I am not sure that the limitation mentioned just before section 3.2 is
>  not so much related to nesting of languages, but rather due to only
>  having a single recursive variable...

The limitation is about nesting. How can it then not be related to nesting?
There is a second limitation that mutual recursion is not possible which is
because we only have a single variable. Perhaps I should make those two more
clear in the paper.

> * I am not entirely convinced by the nu-notation for nullability. The
>  same notation is also used for the greatest fixpoint of a
>  functor. Given that the paper also uses the (standard) mu notation for
>  the least fixpoint, perhaps it might be more clear to write empty (or
>  something along these lines).

The nu notation is quite standard. Brzozowski originally used delta-zero, but I
already use that for a different purpose and I think that would also be
similarly confusing.

> * There is quite some work on (generic programming with) fixpoints of
>  indexed functors or mutually recursive families of datatypes. Citing
>  this as an avenue for further work might help substantiate the
>  argument that these ideas can be extended to handle richer grammars.

Yes, in the discussion on expressiveness we should cite indexed data type
descriptions and how they can be used to model mutually recursive data types.
That work does port over directly to our grammars. However, I did not manage to
finish all the proofs yet, because the indexing makes them quite a bit more
complicated.

> ### Typos
> 
> * book-keeping -> bookkeeping
> 
> * a program which can prove for us -> proves whether or not ...
> 
> * condintional -> conditional
> 
> * 'Most practical programming languages' and 'Many practical programming languages' - repitition.
> 
> * allowed to take the fixpoint of -> of which we are allowed to take the
>  fixpoint (never end a sentence with a proposition).
> 
> * D But (missing punctuation mark)

Bottom of page 11

> * Hocroft -> Hopcroft
> 
> * Danielsson and Anders -> Danielsson
> 
> 
> ----------------
>  REPORT
> ----------------
> 
> This paper formalizes derivative-based parsing for context-free
> grammars with the restriction that there are no mutual recursions
> (i.e., the number of nonterminal is at most one).
> 
> I have a mixed attitude about the paper. On the one hand, since the
> formalization of derivative-based parsing for CFG itself is new, this
> work would be worth to be published in TFP. On the other hand, since
> the technical difficulty of handling derivative-based parsing for
> context-free grammars lies only in the treatment of nonterminals,
> having the restriction on it is a limitation.
> 
> Let me first explain my understanding to assess the limitation. In my
> understanding, the starting point of Might+ 2011 is symbolic
> derivation of both sides of rules, where the derivation of a
> nonterminal is treated as a new nonterminal. For example, consider the
> grammar in p3. From <expr> ::= <expr> + <expr> | 0 | 1 | <stmt>, we
> have
> 
>    d0 <expr> ::= (d0 <expr>) + <expr> | ε | d0 <stmt>
> 
> because we have
> 
>    d0 RHS
>    = d0 (<expr> + <expr> | 0 | 1 | <stmt>)
>    = d0 (<expr> + <expr>) | d0 0 | d0 1 | d0 <stmt>
>    = d0 <expr> + <expr> | ε | ∅ | d0 <stmt>
>    = (d0 <expr>) + <expr> | ε | d0 <stmt>
> 
> Also, from <stmt> ::= <expr> | <stmt> ; <stmt>, we have
> 
>    d0 <stmt> ::= d0 <expr> | (d0 <stmt>) ; <stmt>
> 
> because we have:
> 
>    d0 RHS
>    = d0 (<expr> | <stmt> ; <stmt>)
>    = d0 <expr> | d0 (<stmt> ; <stmt>)
>    = d0 <expr> | (d0 <stmt>) ; <stmt>
> 
> Here, we used the fact that neither <expr> nor <stmt> is
> nullable.
> 
> As this example demonstrates, essentially, the treatment of
> concatenation, choice, empty string, and terminals is nothing
> different from the regular-expression case. Hence, the only difficulty
> lies in the treatment of nonterminals. (Actually, this is highlighted
> in the definition of ⟦_⟧ₒ and its treatment.) As we treat a derivative
> of a nonterminal as a new nonterminal, a derivative of a grammar may
> have twice the number of nonterminals of the original one.
> 
> In a pen-and-paper proof, I expect the correctness of a derivation of
> a grammar would be proved by using the following three steps.
> 
>  1. Correctness of nullability test
> 
>  2. Preservation of the languages of kept nonterminals (such as
>     <expr> and <stml> in the derived grammar)
> 
>  3. The language of a derived nonterminal is indeed a derivative
>     (left quotient) of the language of an original one.
> 
> (Maybe, the Steps 2 and 3 can be proved simultaneously using induction
> on production.) In this paper, Steps 1 and 3 are proved by induction
> on a grammar structure and Step 2 is avoided by having μ.
> 
> I suspect that what will be complicated most by allowing mutual
> recursions (or more than one nonterminal) would be Step 1 because it
> would require at least as many unfoldings as the number of
> nonterminals. In contrast, I may think that Steps 2 and 3 would not
> get so complicated.
> 
> Considering these points, whether I should think of the limitation as
> a severe one or a simplification depends on whether I should think the
> correctness of the nullability test is a core part of the
> derivative-based parsing. I think that there are different standing
> points about this. Some may say that nullability is a common concern
> with CFGs and thus not specific to the derivative-based
> parsing. Others may say that it must be in the core part as the
> derivative-based parsing heavily relies on nullability checks.
> 
> Eventually, I have found the following paper, which involves the
> correctness of the nullability check as a part of their formalization
> (I don't know whether it is the first one).
> 
>    Denis Firsov, Tarmo Uustalu: Certified normalization of context-free
>    grammars, CPP 2015. https://eur03.safelinks.protection.outlook.com/?url=https%3A%2F%2Fdoi.org%2F10.1145%2F2676724.2693177&data=05%7C02%7Cj.s.reinders%40tudelft.nl%7C4020b24233f24d08a73f08dd6c64e624%7C096e524d692940308cd38ab42de0887b%7C0%7C0%7C638785902816936482%7CUnknown%7CTWFpbGZsb3d8eyJFbXB0eU1hcGkiOnRydWUsIlYiOiIwLjAuMDAwMCIsIlAiOiJXaW4zMiIsIkFOIjoiTWFpbCIsIldUIjoyfQ%3D%3D%7C0%7C%7C%7C&sdata=73dHpP77s629c%2FMi%2FEc3bcGrzwXlxt9Ehdq9oIWnTn4%3D&reserved=0
> 
> This means, the correctness of nullability check cannot be a
> contribution by itself. So, I think the authors can state that the
> restriction is a simplification by citing such a paper, clarifying the
> standing point. However, at the same time, I also think that, since we
> can rely on such results for the correctness of the nullability check,
> the authors can support mutual recursion with reasonable effort.
> 
> In conclusion, putting these points in consideration, my evaluation of
> this paper is in-between 0 and 1. I am slightly positive but not that
> positive to score 1.

> To improve the paper's presentation, I suggest
> the authors to consider showing the picture of the formalization:
> especially, what is the core idea of the derivation of context-free
> grammars and what will be the point in proving its correctness.

TODO: We could try to give a better high level overview of the proof.

>  Also,
> please consider citing the above paper to justify your restriction on
> grammars.

TODO: read and cite the paper

> In addition, I would strongly suggest that the author change
> the title, because the present paper is not the first one that shows
> mechanized formalization of context-free grammars.

TODO: With "Type Theoretically" I intended to mean that this is a very specific
perspective on how to formalize, so we don't just present a formalization.

TODO: Consider changing title, e.g. "Towards a Type Theoretic Treatment of Context-Free Languages"

> 
> ## Individual Comments
> 
> p1. "new research continuously emerge". To support this, it would be
> better to cite more recent papers in addition. The latest paper cited
> in this paragraph is published in 2015.
> 
> p2. "<pal> ::= 0 | 1 | 0 <pal> 0 | 1 <pal> 1". Reading the following
> discussions, I suspect that the production of ϵ is missing and the
> correct one should be:
> 
>    <pal> ::= ϵ | 0 | 1 | 0 <pal> 0 | 1 <pal> 1
> 
> p2. "The idea of automatic differentiation." I would feel that the
> idea here would be more similar to symbolic differentiation.
> 
> p3. "... the recursive unfolding we performed ..." However, I don't
> think there is an issue in symbolic treatment, which can derive the
> grammar (d0 <rec>) ::= (d0 <rec>) from <rec> ::= <rec>. Both are valid
> CFGs.
> 
> p3. "Another challenge ... is how to encode their recursive nature." I
> suspect this is still a challenge, because we already have a number of
> existing approaches of mechanized formalization of CFGs. Also, we can
> use a general technique that works for general DSLs, such as de Bruijn
> index and parametric HOAS.
> 
> This is in Haskell, but the following paper discusses CFGs represented
> by using de Bruijn index:
> 
>    Arthur Baars, S. Doaitse Swierstra, Marcos Viera: Typed
>    Transformations of Typed Grammars: The Left Corner Transform.
>    ENTCS 253 (7), 2010.  https://eur03.safelinks.protection.outlook.com/?url=https%3A%2F%2Fdoi.org%2F10.1016%2Fj.entcs.2010.08.031&data=05%7C02%7Cj.s.reinders%40tudelft.nl%7C4020b24233f24d08a73f08dd6c64e624%7C096e524d692940308cd38ab42de0887b%7C0%7C0%7C638785902816949725%7CUnknown%7CTWFpbGZsb3d8eyJFbXB0eU1hcGkiOnRydWUsIlYiOiIwLjAuMDAwMCIsIlAiOiJXaW4zMiIsIkFOIjoiTWFpbCIsIldUIjoyfQ%3D%3D%7C0%7C%7C%7C&sdata=knSvdZ8cwmqrVnAbyGZJFQk4MThzcmaUZ6bV5Q3xWbw%3D&reserved=0
> 
> p6. Explain "Dec.map" and "mk⇔".
> 
> p10. "... we have learned the following three lessons." I suspect that
> at least the first two of the three lessons can be obtained
> straightforwardly from the existing results.
> 
> p15 "His work does cover mutually recursive nonterminals ..." He is
> discussing μ-regular expressions, which is not mutually recursive,
> whereas their languages are the same as the ones of CFGs.
> 
> p15. "... parsing is only the boolean ..." I think that this applies
> also to the proposed approach.
> 
> p15. What are "μ-regular languages"? Anyway, I suspect that at least
> [13] discuss CFGs.
> 
> p16. "... we have formalized (acyclic) context-free grammars ...". In
> my understanding, the proposed approach can handle cyclic ones like
> <rec> ::= <rec>.
> 
> 
> ----------------
>  REPORT
> ----------------
> 
> As it says on the tin, this paper addresses treating context free languages within a type-theoretic framework. Authors extend on existing work in formal languages mechanized in Agda and the line of work lifting derivatives to context-free languages.
> 
> Key technical contributions include defining a suitable fixed-point construction in Agda to safely encode recursive grammars, formalizing language derivatives to parse strings incrementally, and proving correctness of the parsing. Authors identify some present limitations of their approach and indicate those as directions for future work.
> 
> The background section is nice, gentle, readable. The whole thing is well written and explained. This is straight-down-the-middle nice line of work. "Here's some interesting pieces of work by some folk, but they haven't been put together in this way, and there's obviously a missing hole in our knowledge in this area, let me fix it." Overall, this work effectively fills a gap in formal parsing literature.  providing a clean, foundational formalization of context-free language parsing, type theoretically.
> 
> I have only minor notes.
> 
> Forgive me if this is silly, but we should explicitly include the empty string in the grammar; unless we do so, "0110" is not in the language of the given grammar; as written all productions are of odd-length strings.
> 
> The footnote from 2.1 appears on the next page, and that both got me confused looking for where it came from, and also confused as the return type of Parser, making me think "is this 'Set' deliberate here or not?" (and I think it's a mistake)
> 
> 
> * small
> 
> ** "Doing so for the <pal> grammar" -> "Performing this unfolding for the <pal> grammar"
> 
> ** I know it's not what you mean, but what you said is that recursively unfolding the <pal> nonterminal produces the derivative of <pal0> wrt 1.
> 
> "Now, how do we take the derivative of the grammar <pal0> w.r.t. the next bit (1) in our string? [Say explicitly why it can't be done for <pal0> as written.] A simple solution [to the just-mentioned problem] is to recursively unfold the ⟨pal⟩ non-terminal. [After this unfolding, we can take the derivative]."
> 
> ** "see similarity between this type and the"
> 
> Repeat it. "similarity between the type (w : String) → Dec (P w) and". Remove the room for ambiguity.
> 
> ** speficying [specifying] and parsing context-free languages.
> 
> ** to accept it any way [anyway]
> 
> ** if we restrict the [class of] functions
> 
> ** and the[n] we take the derivative of the second
> 

### 2025-04-08 tfp2025 3/review-response.md

> ** from which our disired [desired]
> 
> ** an anonymous reviewer of a draft of this paper have [has]
> 
> ** the interested reader to Hocroft et al. [Hopcroft]
> 
> Use cleveref and you'll never again misspell an author in text.
>

### 2025-04-10 tfp2025 3/src/sections/1-introduction.tex
Address many minor concerns

\begin{code}[hide]%
\>[0]\AgdaSymbol{\{-\#}\AgdaSpace{}%
\AgdaKeyword{OPTIONS}\AgdaSpace{}%
\AgdaPragma{--cubical}\AgdaSpace{}%
\AgdaPragma{--guarded}\AgdaSpace{}%
\AgdaSymbol{\#-\}}\<%
\\
\>[0]\AgdaKeyword{module}\AgdaSpace{}%
\AgdaModule{1-introduction}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
%
\\[\AgdaEmptyExtraSkip]%
\>[0]\AgdaKeyword{open}\AgdaSpace{}%
\AgdaKeyword{import}\AgdaSpace{}%
\AgdaModule{Data.Unit}\<%
\\
\>[0]\<%
\end{code}

@@

@@

>>>>

\begin{code}[hide]%
\>[0]\AgdaKeyword{module}\AgdaSpace{}%
\AgdaModule{1-introduction}\AgdaSpace{}%
\AgdaKeyword{where}\<%
\\
%
\\[\AgdaEmptyExtraSkip]%
\>[0]\AgdaKeyword{open}\AgdaSpace{}%
\AgdaKeyword{import}\AgdaSpace{}%
\AgdaModule{Data.Unit}\<%
\\
\>[0]\<%
\end{code}

@@

Parsing---i.e., the process of recovering structure from strings---is an essential building block for many practical programming applications.
While parsing has been extensively studied, it remains a relevant subject of research where new research questions continuously emerge.
For example, how to compose grammars and parsers~(e.g.,~\cite{SchwerdfegerW09}), dealing with ambiguous parse trees~(e.g.,~\cite{BrabrandGM10,basten-thesis,one-parser-to-rule-them-all}), and parsing grammar formalisms beyond context-free grammars~(e.g.,~\cite{one-parser-to-rule-them-all}).
Research questions such as these serve a practical purpose, but answering them often requires a deep theoretical understanding of the semantics of parsing.

This theoretical understanding can be approached in different ways.
Parsing is often studied using automata theory~\cite{hopcroft-book}.
However, there is value in studying more \emph{denotational} approaches to parsing.
A main purpose of denotational semantics is to abstract away operational concerns, as such concerns tends to be a hindrance for equational reasoning.
Such equational reasoning could be used to study and answer some of the open research questions in the parsers of today and tomorrow.

This paper studies the denotational semantics of parsing for context-free grammars.
While the study is theoretical in nature, the motivation is to provide a foundation for practical future studies on proving the correctness of, e.g., parser optimizations and disambiguation techniques, as well as potentially providing a foundation for building and reasoning about parsers for more expressive grammar formalisms, such as data-dependent grammars~\cite{one-parser-to-rule-them-all}.

We approach the question of giving a denotational semantics of parsing by building on existing work by Elliott~\cite{conal-languages}.
In his recent work, Elliott demonstrated that regular grammars have a simple and direct denotational semantics.
And that we can obtain parsers for such languages that are correct by construction, using \emph{derivatives}.
While it was well-known that we can parse regular grammars using Brzozowski derivatives~\cite{brzozowski}, Elliott's work provides a simple and direct mechanization in Agda's type theory of the denotational semantics of these derivatives.
This mechanization provides an implementation of parsing that is correct by construction, and that we can reason about without relying on (bi-)simulation arguments.
While the parsers obtained in this manner are not exactly performant, the denotational approach opens up the door to exploiting grammar structure to obtain optimized parsers.

Elliott leaves open the question of how the approach scales to more expressive grammar formalisms, such as context-free languages and beyond.
However, the question of using derivatives to parse context-free grammars has been considered by others.
Might et al.~\cite{parsing-with-derivatives} demonstrate how to build parsers from context-free grammars using derivatives and optimizations applied to them, to obtain reasonable performance.
Thiemann's work~\cite{Thiemann17} uses lattice theory and powerset semantics to formalize a notion of partial derivative for a variant of context-free grammars.
In this work, we build on the approach of Elliott and study how to build a simple and direct mechanization in Agda's type theory of the denotational semantics of derivatives for context-free grammars.

The main challenge for our mechanization is the question of how to deal with the recursive nature of context-free languages.

\subsection{The Challenge with Parsing Context-Free Grammars}

We give an overview of what it means to take the derivative of a grammar, how this provides an approach to parsing, and which challenges it poses.

To illustrate, consider the following context-free grammar of palindromic bit strings:
\[
\langle\mathit{pal}\rangle ::= ε \mid 0 \mid 1 \mid 0\, \langle\mathit{pal}\rangle\, 0 \mid 1\, \langle\mathit{pal}\rangle\, 1
\]
Say we want to use this grammar to parse the string 0110.
The idea of parsing with derivatives is to divide the problem into separate steps for each input symbol, each narrowing down the grammar.
After all input symbols have been processed, we can simply check if the final grammar accepts the empty string to see if the whole input is accepted by the original grammar.

We call each of these transformation steps derivatives.
Taking the derivative of a grammar with respect to an input symbol can be done naively by keeping only the productions which start with that symbol and then removing that symbol from the start of each of the productions.
Empty productions (ε) are always removed when taking a derivative.

Using this naive derivation procedure on the $\langle\mathit{pal}\rangle$ grammar with respect to the bit 0 yields the following derived grammar:
\[
\langle\mathit{pal}_0\rangle ::= ε \mid \langle\mathit{pal}\rangle\, 0
\]
This grammar essentially represents the residual parsing obligations after parsing a 0 bit.
The derived grammar contains fewer productions than the original grammar because we have pruned those productions that started with the terminal symbol 1.

For the next derivative step, we encounter a new special case: one of the productions starts with a nonterminal.
A naive solution is to recursively unfold nonterminals until the production starts with a terminal symbol.
For the $\langle\mathit{pal}\rangle$ grammar, we only have to unfold $\langle \mathit{pal} \rangle$ once which yields the following grammar:
\[
\langle\mathit{pal}_0'\rangle ::= ε \mid 0\, 0 \mid 1\, 0 \mid 0\, \langle\mathit{pal}\rangle\, 0\, 0 \mid 1\, \langle\mathit{pal}\rangle\, 1\, 0
\]
By continuing this procedure, with additional recursive unfolding where needed, we eventually yield a grammar that contains the the empty production ε, whereby we can conclude that 0110 is, in fact, a palindromic bit string.

However, naive recursive unfolding does not work for all grammars.
Consider, for example, the infinitely recursive grammar:
\[
\langle\mathit{rec}\rangle ::= \langle\mathit{rec}\rangle
\]
Unfolding this grammar never reveals a starting nonterminal.
While the $\langle\mathit{rec}\rangle$ grammar is contrived, the same issue arise for any \emph{left-recursive} grammar, such as the following grammar of arithmetic expressions (left-recursive because of the $\langle\mathit{expr}\rangle$ non-terminal in the left-most position in the first production):
\[
\langle\mathit{expr}\rangle ::= \langle\mathit{expr}\rangle\, +\, \langle\mathit{expr}\rangle \mid 0 \mid 1
\]

Parsing context-free grammars, thus, requires more sophisticated techniques.

\subsection{Contributions}

This paper tackles the challenges discussed in the previous section by providing a mechanization in Agda of parsing a subset of context-free grammars with derivatives.
The subset of grammars that we consider corresponds to context-free grammars without mutually recursive nonterminals.
In other words, the grammars can consist of multiple nonterminals which may refer to themselves and others, but there may not be a cycle which involves more than one nonterminal.
For example, the following mutually recursive grammar does not fit into the subset of grammars we consider:
\begin{align*}
\langle\mathit{expr}\rangle &::= \langle\mathit{expr}\rangle\, +\, \langle\mathit{expr}\rangle \mid 0 \mid 1 \mid \langle\mathit{stmt}\rangle
\\
\langle\mathit{stmt}\rangle &::= \langle\mathit{expr}\rangle \mid \langle\mathit{stmt}\rangle ; \langle\mathit{stmt}\rangle
\end{align*}
The $\langle\mathit{pal}\rangle$, $\langle\mathit{rec}\rangle$, and $\langle\mathit{expr}\rangle$ grammars from the previous section are all examples of grammars that are in the subset.
We conjecture that the approach we present later can be extended to deal with all context-free grammars, at the cost of some additional bookkeeping.
We leave verifying this conjecture as a challenge for future work.

We make the following technical contributions:
\begin{itemize}
\item We provide a semantics in Agda of context-free grammars without mutual recursion.
\item We provide a derivative-based parser for this class of grammars, along with its simple and direct correctness proof.
\end{itemize}

The paper assumes basic familiarity with Agda.
The rest of this paper is structured as follows.
\Cref{sec:finite-languages} recalls the essential definition from Elliott's work which we subsequently extend in \cref{sec:context-free} to context-free grammars.
\Cref{sec:discussion} discusses expressiveness, performance, and simplicity of our approach, whereas
\cref{sec:related-work} discusses related work, and \cref{sec:conclusion} concludes.

\endinput

@@

%%% Local Variables:
%%% reftex-default-bibliography: ("../references.bib")
%%% End:

### 2025-04-10 tfp2025 3/src/sections/4-discussion.tex
Address many minor concerns

\section{Discussion}

Finally, we want to discuss three aspects of our work: expressiveness, performance, and simplicity.

\jr{TODO: μ-regular expressions have been studied before, cite}
\paragraph{Expressiveness} We conjecture that our grammars which include variables and fixed points can describe any context-free language.
\jr{mention that we only support context-free languages without mutual recursion and how we use a subset of μ-regular languages}.
% We have shown the example of balanced the bracket language which is known to be context-free. Furthermore, Grenrus shows that any context-free grammar can be converted to his grammars \cite{fix-ing-regular-expressions}, which are similar to our grammars. The main problem is showing that mutually recursive nonterminals can be expressed using our simple fixed points, which requires Beki\'c's bisection lemma~\cite{Bekic1984}. Formalizing this in our framework is future work.

Going beyond context-free languages, many practical programming languages cannot be adequately described as context-free languages. For example, features such as associativity, precedence, and indentation sensitivity cannot be expressed directly using context-free grammars. Recent work by Afroozeh and Izmaylova~\cite{one-parser-to-rule-them-all} shows that all these advanced features can be supported if we extend our grammars with data-dependencies. Our framework can form a foundation for such extensions and we consider formalizing it as future work.

\paragraph{Performance}
\jr{cite Jeremy Yallop's work}
For a parser to practically useful, it must at least have linear asymptotic complexity for practical grammars. Might et al. \cite{parsing-with-derivatives} show that naively parsing using derivatives does not achieve that bound, but optimizations might make it possible. In particular, they argue that we could achieve $O(n|G|)$ time complexity (where $|G|$ is the grammar size) if the grammar size stays approximately constant after every derivative. By compacting the grammar, they conjecture it is possible to achieve this bound for any unambiguous grammar. We want to investigate if similar optimizations could be applied to our parser and if we can prove that we achieve this bound.

\paragraph{Simplicity}
One of the main contributions of Elliott's type theoretic formalization of languages~\cite{conal-languages} is its simplicity of implementation and proof. To be able to extend his approach to context-free languages we have had to introduce some complications.
\jr{TODO: finish this paragraph}
% Most notably, we use fuel to define the semantics of our grammars. We have explored other approaches such as using guarded type theory, but we did not manage to significantly simplify our formalization. Furtheremore, we expect that many proofs remain simple despite our fuel-based approach.

>>>>

\section{Discussion}\label{sec:discussion}

Zooming back out, we want to discuss three aspects of our work: expressiveness, performance, and simplicity.

\paragraph{Expressiveness} Our language descriptions are similar to μ-regular
expressions, which have been shown to be equivalent to context-free
grammars~\cite{Thiemann17}. The major difference is that we only allow variables
to refer to the nearest enclosing fixed point and not back to earlier fixed
points. Each fixed point roughly corresponds to a nonterminal in the
context-free grammar, so we conjecture this limitation means we only support
context-free grammars without mutual recursion. We hope to expand our work to cover mutually recursive context-free grammars in future work.

Furthermore, an anonymous reviewer of a draft of this paper has pointed out that
simply showing an equivalence of expressive power does not mean μ-regular
expressions are as easy to use as context-free grammars in practice. In future
work, we hope to explore this further by specifying and parsing more
representative and practical languages.

Going beyond context-free languages, many practical programming languages cannot
be adequately described as context-free languages. For example, features such as
associativity, precedence, and indentation sensitivity cannot be expressed
directly using context-free grammars. Recent work by Afroozeh and
Izmaylova~\cite{one-parser-to-rule-them-all} shows that all these advanced
features can be supported if we extend our grammars with data-dependencies. In
future work, we want to explore using our framework as a foundation for such
extensions.

\paragraph{Performance} For a parser to practically useful, it must at least have linear asymptotic complexity for practical grammars. Might et al.~\cite{parsing-with-derivatives} show that naively parsing using derivatives does not achieve that bound, but optimizations might make it possible. In particular, they argue that we could achieve $O(n|G|)$ time complexity (where $|G|$ is the grammar size) if the grammar size stays approximately constant after every derivative. By compacting the grammar, they conjecture it is possible to achieve this bound for all unambiguous grammars. In future work, we want to investigate if similar optimizations could be applied to our parser and if we can prove that we achieve this bound.

\paragraph{Simplicity} One of the main contributions of Elliott's type theoretic formalization of languages~\cite{conal-languages} is its simplicity of implementation and proof. To be able to extend his approach to context-free languages we have had to introduce language descriptors, which are a slight complication. For example, descriptors made it harder to parameterize the symbolic representation of a language with its semantics as Elliott does in his paper, allowing him to write a correct by construction derivative function. Instead, we have split the derivative function into a syntactic transformation and an external correctness proof. Nevertheless, we think our approach retains much of the simplicity of Elliott's work.
